---
title: "BIOST 515-Project"
author: "Michael Cork, Claire Rothschild,  Baohai WattShao, Aji John"
date: "Winter Quarter 2018"
output:
  pdf_document: default
---


We are interested in changes in plasma lipid biomarkers for coronary heart disease (CHD) after hormone replacementtherapy in a sample of 2,763 womenfrom the Heart and Estrogen/progestin Replacement Study (HERS). HERS wasa randomized, double-blind, placebo-controlled trial designed to test the efficacy and safety of estrogen plus progestin therapy for prevention of recurrent coronary heart disease (CHD) events in women.  

```{r setup, include=FALSE}
### Setting up the packages, options we'll need:
library(knitr)
library(ggplot2)
library(tidyverse)
library(uwIntroStats)
library(gridExtra)
knitr::opts_chunk$set(echo = FALSE)

```


    
# Questions of interest to be answered
The data tobe analyzed for this projectis a subset of the data collected from the 2,763 women in the Heart and Estrogen/progestin Replacement Study (HERS) clinical trial of hormone therapy.The questions to be addressed are:   
```{r ,echo=FALSE,include=FALSE}
### Load data
HERS <- read.csv("data/HERSdatasub.csv",header=TRUE) 
HERS$D.HDL <-  round(HERS$HDL.1 - HERS$HDL)
HERS$D.LDL <-  round(HERS$LDL.1 - HERS$LDL)
HERS$D.TG <-  round(HERS$TG.1 - HERS$TG)

```



## 1.
What associations exist between the plasma lipid biomarkers for CHD at baseline (i.e., prior to randomized treatment assignment) and the available data on participant demographics (age, race, BMI), behavior (smoking, alcohol consumption, physical activity), and available clinical and laboratory measures of organ system functioning (e.g, glucose, blood pressure)? 

univariate

## 2.
Is there any evidence of hormone therapy treatment effects on the plasma lipid biomarkers after one year of treatment?  

### Profile of subjects studied

```{r,echo=FALSE,include=TRUE}

HERS%>% filter(complete.cases(HERS)) %>% group_by(HT,age) %>% 
  ggplot(aes(age, fill = HT)) + geom_histogram(binwidth = 1) + theme_classic()

```

### BMI Profile of subjects studied by Control/Treatment

```{r ,echo=FALSE,include=TRUE}
### Is there any evidence HERS
kable(HERS%>%filter(complete.cases(HERS)) %>% 
        group_by(HT) %>%
        summarise(n=n(),sdBMI=sd(BMI),meanBMI=mean(BMI),sdBMI.1=sd(BMI.1),meanBMI.1=mean(BMI.1))%>%
        select(c('HT','n','meanBMI','sdBMI','meanBMI.1','sdBMI.1')) %>%
        as.data.frame() , format = "markdown")
 
## HERS%>%filter(complete.cases(HERS)) %>% nrow()
##[1] 2580
## HERS%>%nrow()
## 2763

```
```{r ,echo=FALSE,include=TRUE}
### Is there any evidence HERS
kable(HERS%>%filter(complete.cases(HERS)) %>% 
        group_by(HT) %>%
        summarise(n=n(),sdLDL=sd(LDL),meanLDL=mean(LDL),sdLDL.1=sd(LDL.1),meanLDL.1=mean(LDL.1))%>%
        select(c('HT','n','meanLDL','sdLDL','meanLDL.1','sdLDL.1')) %>%
        as.data.frame() , format = "markdown")
 
## HERS%>%filter(complete.cases(HERS)) %>% nrow()
##[1] 2580
## HERS%>%nrow()
## 2763

```


```{r,echo=FALSE,include=TRUE}


```

## 3.
Is there any attenuation of the hormone therapy treatment effects after adjustment for known risk factors for CHD? 



## 4.
Do any of the hormone therapy treatment effects identified above for the plasma lipid biomarkers differ according to race/ethnicity, statin medication use,smoking behavior, or alcohol consumption.

### Show the distribution by biomarkers between Placebo and Treatment 

```{r,echo=FALSE,include=TRUE}
db1<- HERS%>% filter(complete.cases(HERS)) %>% group_by(HT,age) %>% 
  ggplot(aes(age,D.HDL, color = HT)) + geom_point() + 
  stat_smooth(size=1.5, method = "loess", level = 0.95, fullrange = TRUE, se = FALSE) +
  theme_classic() + labs(colour = "")
db2<- HERS%>% filter(complete.cases(HERS)) %>% group_by(HT,age) %>% 
  ggplot(aes(age,D.LDL, color = HT)) + geom_point() + 
  stat_smooth(size=1.5, method = "loess", level = 0.95, fullrange = TRUE, se = FALSE) +
  theme_classic() + labs(colour = "")
db3<- HERS%>% filter(complete.cases(HERS)) %>% group_by(HT,age) %>% 
  ggplot(aes(age,D.TG, color = HT)) + geom_point() + 
  stat_smooth( method = "loess") +
  theme_classic() + labs(colour = "")
grid.arrange(db1, db2, db3,nrow=3)

```



### Checking by Race

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}




g1 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.LDL)) + geom_boxplot() +
  ggtitle("LDL.1") +  stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity") + ggtitle("Differ Race/Ethnicity ?")

g2 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.HDL)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity") + ggtitle(" ")

g3 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity")  + ggtitle(" ")

grid.arrange(g1, g2, g3,ncol=3)
```

By looking at the boxplots(SE-Max, Mean, SE-Min), we could hypothesize that TG might be affected by race. We could do an ANOVA to confirm that.

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}
oned_alt<- glm(formula = D.LDL~  raceth +age + smoking + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])
oned.null<- glm(formula = D.LDL~  age + smoking + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])

anova(oned_alt,oned.null,test = "Chisq")
```

For LDL, the effect is found to be not significant(p-value 0.96), but for TG and HDL, the effect do differ by race/ethnicity(p-value 0.01, and p-value < 0.001 respectively). The analysis was facilitated by ANOVA where null model included demograhics(excluding BMI),diabetic indicator and behavior. Race/Ethnicity was omitted in the null model.

### Checking by Statin Use

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}

g1 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.LDL)) + geom_boxplot() +
  stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use")  + ggtitle("Differ by Statin use ?") 

g2 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.HDL)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use") 

g3 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use")  

grid.arrange(g1, g2, g3,ncol=3)
```

By looking at the boxplots(SE-Max, Mean, SE-Min), we could hypothesize that LDL might be affected by statin use We could do an ANOVA to confirm that.

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}
oned_alt<- glm(formula = D.LDL ~  raceth +age + smoking + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])
oned.null<- glm(formula = D.LDL ~  raceth +age + smoking + drinkany+exercise+diabetes, data = HERS[complete.cases(HERS),])

anova(oned_alt,oned.null,test = "Chisq")
```
For TG,and HDL the effect is found to be not significant(p-value = 0.21, and p-value =0.97 respectively), but for LDL, the effect do differ by statin use(p-value < 0.001). The analysis was facilitated by ANOVA where null model included demograhics(excluding BMI),diabetic indicator and behavior. Statin use was omitted in the null model.

### Smoking Behavior

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}

g1 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.LDL)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking") + ggtitle("Differ by Smoking ?") 

g2 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.HDL)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking") 

g3 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.TG)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking") 

grid.arrange(g1, g2, g3,ncol=3)
```

By looking at the boxplots(SE-Max, Mean, SE-Min), we could not really hypothesize the affects by smoking behavior. We could do an ANOVA to confirm which ones are significant

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}
oned_alt<- glm(formula = D.TG ~  raceth +age + smoking + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])
oned.null<- glm(formula = D.TG ~  raceth +age  + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])

anova(oned_alt,oned.null,test = "Chisq")
```
For TG,LDL,HDL the effect is found to be not significant(p-value = 0.29, p-value = 0.42, and p-value = 0.99 respectively). The analysis was facilitated by ANOVA where null model included demograhics(excluding BMI),diabetic indicator and behavior. Smoking indicator was omitted in the null model.

### Alcohol use

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}

g1 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.LDL)) + geom_boxplot() +
  stat_summary(
  colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol")  +ggtitle("Differ by Alcohol use?") 

g2 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.HDL)) + geom_boxplot() +
 stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol") 

g3 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol") 

grid.arrange(g1, g2, g3,ncol=3)
```

By looking at the boxplots(SE-Max, Mean, SE-Min), we could not really hypothesize the affects by alcohol use. We could do an ANOVA to confirm which ones are significant

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}
oned_alt<- glm(formula = D.TG ~  raceth +age + smoking + drinkany+exercise+statins+diabetes, data = HERS[complete.cases(HERS),])
oned.null<- glm(formula = D.TG ~  raceth +age + smoking +exercise+statins+diabetes, data = HERS[complete.cases(HERS),])

anova(oned_alt,oned.null,test = "Chisq")
```
For TG,LDL,HDL the effect is found to be not significant(p-value = 0.99, p-value = 0.13, and p-value = 0.24 respectively). The analysis was facilitated by ANOVA where null model included demograhics(excluding BMI),diabetic indicator and behavior. Drinking indicator was omitted in the null model.

```{r,echo=FALSE,warning=FALSE,error=FALSE,message=FALSE,include=TRUE}

g1 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.LDL)) + geom_boxplot() +
  ggtitle("LDL.1") +  stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity") 

g2 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.HDL)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity") + ggtitle(" ")

g3 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(raceth), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Race/Ethnicity")  + ggtitle(" ")

g4 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.LDL)) + geom_boxplot() +
  stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use")  

g5 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.HDL)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use") 

g6 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(statins), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Statin Use")  

g7 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.LDL)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking")

g8 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.HDL)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking") 

g9 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(smoking), D.TG)) + geom_boxplot() +
   stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Smoking") 

g10 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.LDL)) + geom_boxplot() +
  stat_summary(
  colour = "red",size=0.5) + theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol")  

g11 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.HDL)) + geom_boxplot() +
 stat_summary(colour = "red",size=0.5) + theme_classic() +theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol") 

g12 <- HERS%>% filter(complete.cases(HERS)) %>% filter(HT %in%  c('hormone therapy')) %>% ggplot( aes(factor(drinkany), D.TG)) + geom_boxplot() +
    stat_summary(colour = "red",size=0.5) +theme_classic() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) +     xlab("Alcohol") 

grid.arrange(g1, g2, g3,g4, g5, g6,g7, g8, g9,g10, g11, g12,ncol=3)
```